---
title: "Mimic-Fasting Diet Plan"
author: "Sandra Kilpatrick"
date: "August 05, 2015"
output: html_document
---
#Does Mimic Fasting have health benefits?

I think those of us who work in downtown Chattanooga take for granted how much walking we do every day.  I would walk from my car to the office.  Sometimes, I forgot something and made that trek again, then a meeting on another floor, walk to Chattz for coffee mid-morning, then lunch somewhere near the aquarium, more meetings, back to the car at the end of the day. So when I stopped going into the office and didn't walk as much, I was very surprised at the effects on my body.  Since I was going through the Coursera data science program, I was spending long hours in front of my computer and trying very hard to remember to get up and walk around. I have put on a few pounds, and I'm not happy about it. I was struggling to get up and walk enough, and the pounds kept piling on.

I ran across an article about a new diet plan that got my attention. It's called the "Mimic Fasting Diet." (http://www.telegraph.co.uk/news/science/science-news/11683736/Five-day-fasting-diet-slows-down-ageing-and-may-add-years-to-life.html) I really like the fact that the diet is low-protein.  I know high-protein is all the rage, but my husband has polycystic kidneys so we avoid lots of protein in order to help protect his kidneys.

Apparently, fasting has a lot of benefits. In addition to losing weight, fasting can slow the aging process and  reduce the risk of some age-related diseases like cancer and diabetes.

So this is the structure of the diet plan:
Reduce calorie intake for five days each month and eat food that is comprised of:

*Day 1 - 1090 calories
*Protein - 10% of the calories
*Fat - 56% calories
*Carbs - 34% calories

*Days 2 - 5 - 725 calories
*Protein - 9%
*Fat - 44%
*Carbs - 47%

There aren't very many, if any, meal plans available for this diet, so I decided to create my own. I pulled the USDA Nutrition database from the USDA web site (http://ndb.nal.usda.gov/ndb/doc/index).  The following is the code, written in R, that I have so far to join the datasets and do some analysis.

To access the files and r code, access my github repository:  (https://github.com/SandraKilpatrick/MimicFastingDiet.git)

#R code below
Set directory and load libraries
```{r echo=FALSE, eval=TRUE}
rm(list=ls())
setwd("~/DataScience/NutritionDatabase")

library(dplyr)
library(httr)
library(reshape)
library(reshape2)
library(ggplot2)
library(tidyr)
library(extrafont)
```
Read files into R
```{r echo=TRUE, cache=TRUE}
parse_file <- function(x) {
  base_url <- "http://www.ars.usda.gov/SP2UserFiles/Place/12354500/Data/SR27/asc/"
  resp <- GET(paste0(base_url, x))
  
  text <- content(resp, "text", encoding = "ISO-8859-1")
  read.delim(text = text, sep = "^", quote = "~", na.strings = c("^^", "~~"),
             header = FALSE, stringsAsFactors = FALSE) %>% tbl_df()
 
}
```
Create food
```{r echo=TRUE, cache=TRUE}
food <- parse_file("FOOD_DES.txt")
names(food) <- c("NDB.No", "FdGrp.Cd", "Long.Desc", "Shrt.Desc", "ComName", "ManufacName","Survey", "Ref.desc", "Refuse", "SciName", "N.Factor", "Pro.Factor", "Fat.Factor","CHO.factor")
food$NDB.No<-as.character(food$NDB.No)
food$FdGrp.Cd<-as.character(food$FdGrp.Cd)
food$Survey <- food$Survey == "Y"
```
Create food group
```{r echo=TRUE, cache=TRUE}
food.group <- parse_file("FD_GROUP.txt")
names(food.group) <- c("FdGrp.Cd","FdGrp.Desc")
food.group$FdGrp.Cd<-as.character(food.group$FdGrp.Cd)
```
Create nutrient
```{r echo=TRUE, cache=TRUE}
nutrient <- parse_file("NUT_DATA.txt")
names(nutrient) <- c("NDB.No", "Nutr.No", "Nutr.Val", "Num.Data.Pts", "Std.Err","Src.Cd", "Deriv.Cd", "Ref.NDB.No", "Add.Nutr.Mark", "Num.Studies", "Min","Max", "DF", "Low.EB", "Up.EB", "Stat.cmt", "AddMod.Date", "CC")
nutrient$Add.Nutr.Mark[nutrient$Add.Nutr.Mark == ""] <- NA
nutrient$NDB.No <- as.character(nutrient$NDB.No)
nutrient$Nutr.No <- as.character(nutrient$Nutr.No)
```
Create nutrient definition file
```{r echo=TRUE, cache=TRUE}
nutrient.def <- parse_file("NUTR_DEF.txt")
names(nutrient.def) <- c("Nutr.No","Units","Tagname","NutrDesc","Num.Dec","SR.Order")
nutrient.def$Nutr.No <- as.character(nutrient.def$Nutr.No)
```
Create weight file
```{r echo=TRUE, cache=TRUE}
weight <- parse_file("WEIGHT.txt")
names(weight) <- c("NDB.No","Seq","Amount","Msre.Desc","Gm.Wgt","Num.Data.Pts","Std.Dev")
weight$NDB.No <- as.character(weight$NDB.No)
weight$Seq<-as.character(weight$Seq)
```
Explore the dataset
```{r echo=TRUE}
summary(food)
str(food)
```
Join the datasets and select the variables to analyze
```{r echo=FALSE, cache=TRUE}
#Join food, food group, nutrient, nutrient.def and weight datasets
mimicfast <- food %>%
    left_join(food.group) %>%
    select(NDB.No,FdGrp.Cd,Shrt.Desc,FdGrp.Desc) %>%
    left_join(nutrient) %>%
    select(NDB.No,FdGrp.Cd,Shrt.Desc,FdGrp.Desc,Nutr.No,Nutr.Val) %>%
    left_join(nutrient.def) %>%
    select(NDB.No,FdGrp.Cd,Shrt.Desc,FdGrp.Desc,Nutr.No,Nutr.Val,Units,NutrDesc) %>%
    full_join(weight) %>%
    select(food.id=NDB.No,foodgrp.id=FdGrp.Cd,description=Shrt.Desc,food.group=FdGrp.Desc,nutr.id=Nutr.No,nutr.val=Nutr.Val,Units,nutrient=NutrDesc,seq=Seq,amount=Amount,measure.desc=Msre.Desc,grams=Gm.Wgt)
```
Create the dataset for day 1 with variables we need
```{r echo=TRUE, cache=TRUE}
mimicfast.a <- mimicfast %>%
  #Filter food groups we don't care about
  filter(!grepl("Baby Foods|American Indian/Alaska Native Foods|American Indian/Alaska Native Foods|Fast Foods|Restaurant Foods",food.group))%>%
  #filter for only the nutrients we are exploring
  filter(nutr.id==203|nutr.id==204|nutr.id==205|nutr.id==208)%>%
  arrange(food.id,nutr.id) %>%
  #create new variables
  mutate(uniq.id=paste(food.id,seq)) %>%
  mutate(nutrients.per.gram=nutr.val/100,nutrients.per.serving=nutrients.per.gram*grams) %>%
  arrange(food.id,nutr.id)

head(mimicfast.a)
```
Establish nutrient objectives
```{r echo=TRUE, cache=TRUE}

#Day 1 variables
calorie.obj.day.1<-1090
protein.obj.day<-.10
fat.obj.day<-.56
carb.obj.day<-.34

protein.allowance.per.day.1<-calorie.obj.day.1*protein.obj.day
fat.allowance.per.day.1<-calorie.obj.day.1*fat.obj.day
carb.allowance.per.day.1<-calorie.obj.day.1*carb.obj.day

calorie.allowance.per.serving.day.1<-calorie.obj.day.1/3
protein.allowance.per.serving.day.1<-protein.allowance.per.day.1/2
fat.allowance.per.serving.day.1<-fat.allowance.per.day.1/3
carb.allowance.per.serving.day.1<-carb.allowance.per.day.1/3

#Day 2 variables
calorie.obj.day.2<-725
protein.obj.day<-.09
fat.obj.day<-.44
carb.obj.day<-.47

protein.allowance.per.day.2<-calorie.obj.day.1*protein.obj.day
fat.allowance.per.day.2<-calorie.obj.day.2*fat.obj.day
carb.allowance.per.day.2<-calorie.obj.day.2*carb.obj.day

calorie.allowance.per.serving.day.2<-calorie.obj.day.2/3
protein.allowance.per.serving.day.2<-protein.allowance.per.day.2/2
fat.allowance.per.serving.day.2<-fat.allowance.per.day.2/3
carb.allowance.per.serving.day.2<-carb.allowance.per.day.2/3
```
Calculate difference in nutrient servings
```{r echo=TRUE, cache=TRUE}
#Calorie per serving variance - Day 1
calories.day.1 <- mimicfast.a %>%
  filter(nutr.id==208) %>%
  mutate(calorie.var.day.1=abs(calorie.allowance.per.serving.day.1-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
protein.day.1 <- mimicfast.a %>%
  filter(nutr.id==203) %>%
  mutate(protein.var.day.1=abs(protein.allowance.per.serving.day.1-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
fat.day.1 <- mimicfast.a %>%
  filter(nutr.id==204) %>%
  mutate(fat.var.day.1=abs(fat.allowance.per.serving.day.1-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
carb.day.1 <- mimicfast.a %>%
  filter(nutr.id==205) %>%
  mutate(carb.var.day.1=abs(carb.allowance.per.serving.day.1-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)

#Calorie per serving variance - Day 2
calories.day.2 <- mimicfast.a %>%
  filter(nutr.id==208) %>%
  mutate(calorie.var.day.2=abs(calorie.allowance.per.serving.day.2-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
protein.day.2 <- mimicfast.a %>%
  filter(nutr.id==203) %>%
  mutate(protein.var.day.2=abs(protein.allowance.per.serving.day.2-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
fat.day.2 <- mimicfast.a %>%
  filter(nutr.id==204) %>%
  mutate(fat.var.day.2=abs(fat.allowance.per.serving.day.2-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
carb.day.2 <- mimicfast.a %>%
  filter(nutr.id==205) %>%
  mutate(carb.var.day.2=abs(carb.allowance.per.serving.day.2-nutrients.per.serving)) %>%
  arrange(food.id,nutr.id)
```
Join the Day 1 datasets
```{r echo=FALSE, cache=TRUE}
nutrients.var.day.1 <- mimicfast.a %>%
  left_join(calories.day.1)%>%
  left_join(protein.day.1) %>%
  left_join(fat.day.1) %>%
  left_join(carb.day.1) %>%
  arrange(food.id,nutr.id)

head(nutrients.var.day.1)
```
Transform data set from wide to long
```{r echo=FALSE, cache=TRUE}
mimicfast.data.long.1<-melt(data.frame(nutrients.var.day.1),
id.vars=c("food.id","foodgrp.id","description","food.group","nutr.id","nutrient","measure.desc","grams","uniq.id","nutrients.per.gram","nutrients.per.serving"),
measure.vars=c("calorie.var.day.1","protein.var.day.1","fat.var.day.1","carb.var.day.1"),
variable.name=c("nutrient.allow.meal.desc.day.1"),
value.name="nutrient.allow.meal.value.day.1",na.rm=TRUE
)

mimicfastdiet.1<-select(mimicfast.data.long.1,food.id,foodgrp.id,description,food.group,nutr.id,nutrient,measure.desc,grams,uniq.id,nutrients.per.gram,nutrients.per.serving,nutr.day.1=variable,nutr.var.day.1=value)

head(mimicfastdiet.1)

```
Join the Day 2 datasets
```{r echo=FALSE, cache=TRUE}
nutrients.var.day.2 <- mimicfast.a %>%
  left_join(calories.day.2) %>%
  left_join(protein.day.2) %>%
  left_join(fat.day.2) %>%
  left_join(carb.day.2) %>%
  arrange(food.id,nutr.id)

head(nutrients.var.day.2)
```
Data set from wide to long
```{r echo=TRUE, cache=TRUE}
mimicfast.data.long.2<-melt(data.frame(nutrients.var.day.2),
id.vars=c("food.id","foodgrp.id","description","food.group","nutr.id","nutrient","measure.desc","grams","uniq.id","nutrients.per.gram","nutrients.per.serving"),
measure.vars=c("calorie.var.day.2","protein.var.day.2","fat.var.day.2","carb.var.day.2"),
variable.name="nutrient.allow.meal.desc.day.2",
value.name="nutrient.allow.meal.value.day.2",na.rm=TRUE
)
mimicfastdiet.2<-select(mimicfast.data.long.2,food.id,foodgrp.id,description,food.group,nutr.id,nutrient,measure.desc,grams,uniq.id,nutrients.per.gram,nutrients.per.serving,nutr.day.2=variable,nutr.var.day.2=value)

head(mimicfastdiet.2)
```

Join datasets for both days
```{r echo=TRUE, cache=TRUE}
mimicfastdiet<- (data.frame(mimicfastdiet.1) %>%
  left_join(mimicfastdiet.2)) %>%
  arrange(food.id,nutr.id)

head(mimicfastdiet)
```

Write data set to file
```{r echo=TRUE, cache=TRUE}
write.table(mimicfastdiet,"~/DataScience/NutritionDatabase/mimicfastdiet.txt",sep="\t")
```

#Graphs that aren't useful yet

Bar Graph
```{r}
ggplot(data=mimicfastdiet,aes(x=food.group,y=nutrients.per.serving))+
  geom_bar(stat="identity")

```

Another Bar Graph
```{r echo=TRUE}
ggplot(data=mimicfastdiet, aes(x=food.group, y=nutrients.per.serving, fill=nutrients.per.gram)) +
    geom_bar(stat="identity")
```

Another Bar Graph
```{r echo=TRUE}
ggplot(data=mimicfastdiet, aes(x=food.group)) +
    geom_bar(stat="bin")

```

Line chart
```{r echo=TRUE}
ggplot(data=mimicfastdiet, aes(x=food.group, y=nutrients.per.serving, group=1)) +
    geom_line()

```

Another Line Chart
```{r echo=TRUE}
ggplot(data=mimicfastdiet, aes(x=food.group, y=nutrients.per.serving, group=1)) +
    geom_line() +
    geom_point()

```

Another Line Chart
```{r echo=TRUE}
ggplot(data=mimicfastdiet, aes(x=food.group, y=nutrients.per.serving, group=1)) + 
    geom_line(colour="red", linetype="dashed", size=1.5) + 
    geom_point(colour="red", size=4, shape=21, fill="white")

```

Scatter and Line Plot
```{r}
ggplot(data=mimicfastdiet, aes(x=food.group, y=nutrients.per.serving, group=1)) +
    geom_line() +
    geom_point() +
    expand_limits(y=0) +
    xlab("Food Group") + ylab("Nutrients per Serving") +
    ggtitle("Mimic Fasting Diet")
```

I haven't finished the charts nor have I completed the coding to optimize the food to the set standards, so have some fun with the code.

This is the current code file if you want to look at the dataset.  Feel free to email me at sandra.kilpatrick@chattanoogadatascience.com if you have questions over the weekend or prior to Tuesday's meeting.